---
title: "Untitled"
author: "Kiarash Rastegar"
date: "3/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list=ls())
library(car)
library(psych)
library(ggplot2)
```


# read in and centering data
```{r}
gut = read.csv("gut_diversity.csv")
gut$age = scale(gut$age)
gut$weight = scale(gut$weight)
gut$exercise = scale(gut$exercise)

```

# Building Linear Model 
```{r}

mod1 = lm(diversity~age*exercise, data = gut)
mod1b = lm(diversity~exercise*age, data = gut)
summary(mod1)

Anova(mod1, type=3, test = "F")
summary(mod1b)
Anova(mod1b, type=3, test = "F")

pairs.panels(gut[,c("age", "exercise")])# it looks like age and exercise have no correlation with each other 

# I created mod1b to see if the data is balanced and ran the summary function on it to see if it gave the same result. Since it gave the same results as mod1 summary, I concluded that the data is balanced. 

```

# Visualizing the relationship between diversity and age, as well as, diversity and exercise separately 
```{r}

ggplot(gut, aes(x = age, y=diversity)) + 
  geom_point() + 
  theme_classic()+
  ggtitle("Diversity vs Age")+
  geom_smooth(method = "lm")
  
ggplot(gut, aes(x = exercise, y=diversity)) + 
  geom_point() + 
  theme_classic()+
  ggtitle("Diversity vs Exercise")+
  geom_smooth(method = "lm")

```

Without looking at the summary statistics we can see from our graphs that exercise and gut diversity seem to be correlated. As we increase the amount of exercise we also see a direct increase in the amount of gut diversity. When looking at the graph of diversity and age we do not see this same drastic correlation. To determine the relationship of gut diversity with age and exercise I will plot a 3d model to determine there correlation.

# 3D plot
```{r}
library(plotly)

# set the x, y, and z variables
x <- as.numeric(gut$age) # have to change these values to numeric so that they can be compatible data type with lm 
y <- as.numeric(gut$exercise)
z <- as.numeric(gut$diversity)

# Compute the linear regression 
fit <- lm(z ~ x*y)

# create a grid from the x and y values (min to max) and predict values for every point
# this will become the regression plane
grid.lines = 40
x.pred <- seq(min(x), max(x), length.out = grid.lines)
y.pred <- seq(min(y), max(y), length.out = grid.lines)
xy <- expand.grid( x = x.pred, y = y.pred)
z.pred <- matrix(predict(fit, newdata = xy), 
                 nrow = grid.lines, ncol = grid.lines)

# making variables for my axis titles
x.ax = list(title="Age")
y.ax = list(title="Exercise")
z.ax = list(title="Gut Diversity")

# scatter plot with regression plane
plot_ly(color = I("black"), showlegend = F) %>%
  add_markers(data = gut, z = z, x = x, y = x, opacity = 0.5) %>%
  add_surface(x = ~x.pred, y = ~y.pred, z = ~z.pred, showscale = F, opacity = 0.8) %>%
  add_paths(data = xy,x = ~x.pred, y = ~y.pred, z = ~z.pred ) %>%
  layout(title = "3D Plot of LM", scene = list(xaxis=x.ax,yaxis=y.ax ,zaxis=z.ax))

```

As we can see from the 3D graph, age gives a slight contribution to gut diversity (z-axis), but the height of the plane is dominated by the effect of exercise on gut diversity. This 3D graph supports the observations that I made when looking at the individual 2D graphs, which are shown in the previous chunk. 